---
title: "Week 21: Energy Consumption"
description: "Exploring global energy consumption trends using EIA data."
author: "Daniel B. Mason (Cooper)"
date: 2026-05-26
date-modified: last-modified
categories: [energy, environment]
freeze: true
format:
  html:
    toc: true
    toc-location: left
    code-fold: show
    code-tools: true
---

**Data Source**: [TidyTuesday 2026 Data](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026)

```{r}
#| label: setup
#| message: false
source("renv/activate.R")

library(dplyr)
library(ggplot2)
library(tidyr)
library(tidytuesdayR)

source("../utils/themes.R")
source("../utils/palette.R")
source("../utils/helpers.R")

## load data using tidytuesdayR helper function
tt <- tidytuesdayR::tt_load(2026, week = 21)

```

```{r}
#| label: explore data

df <- tt$energy_cleaned

# Dataset dimensions and year range
cat("Rows:", nrow(df), "| Cols:", ncol(df), "\n")
cat("Year range:", min(df$yr), "–", max(df$yr), "\n")
cat("Countries:", n_distinct(df$country_name), "\n")

# Check missingness on key columns
df |>
  select(
    country_name, yr,
    renewable_energy_consumption_tfec_pct,
    access_electricity_total_pop_pct,
    total_final_consumption_terajoules
  ) |>
  summarise(across(everything(), \(x) sum(is.na(x))))

```

```{r}
#| label: wrangle data

library(stringr)

focus_countries <- c("United States", "China", "Germany", "India", "Brazil")

renewable_trends <- df |>
  filter(country_name %in% focus_countries) |>
  select(
    country_name,
    yr,
    renewable_energy_consumption_tfec_pct,
    traditional_biomass_consumption_tfec_pct,
    access_electricity_total_pop_pct
  ) |>
  # Separate modern renewables (total renewable minus traditional biomass)
  mutate(
    modern_renewable_pct = renewable_energy_consumption_tfec_pct -
      coalesce(traditional_biomass_consumption_tfec_pct, 0)
  ) |>
  filter(!is.na(renewable_energy_consumption_tfec_pct))

```

```{r}
#| label: visualize 

country_colors <- c(
  "Brazil"        = tt_colors[["blue"]],
  "China"         = tt_colors[["vermillion"]],
  "Germany"       = tt_colors[["green"]],
  "India"         = tt_colors[["orange"]],
  "United States" = tt_colors[["sky_blue"]]
)

main_plot <- ggplot(
  renewable_trends,
  aes(x = yr, y = renewable_energy_consumption_tfec_pct, color = country_name)
) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = country_colors) +
  labs(
    title    = "Renewable Energy as Share of Total Final Energy Consumption",
    subtitle = "Five major economies, 1990–2010",
    caption  = "Source: TidyTuesday W21 2026 · EIA / World Bank | @cooperdbm",
    x        = NULL,
    y        = "Renewable Share of TFEC (%)",
    color    = NULL
  ) +
  theme_tt()



```

## Visualization

```{r}
#| label: main-plot

ggplot2::ggsave(
  "main_plot.png",
  plot   = main_plot,
  width  = 9,
  height = 5.5,
  dpi    = 300
)

main_plot
```

## Key Findings

- **Brazil and India** began the period with high renewable shares (~50% and ~57% respectively), but both declined — driven by growing fossil fuel consumption as their economies expanded.

- **China** showed a similar downward trend, dropping from ~33% to ~19% as rapid industrialization increased total energy demand faster than renewable capacity.

- **Germany and the United States** started low (~2–6%) but grew modestly through the 2000s, reflecting early wind and solar policy investments; Germany's growth was notably faster.

- Across all five countries, high renewable shares often reflect reliance on traditional biomass rather than modern clean energy — an important distinction for interpreting "renewable" in developing economies.

## Methodology

The dataset (`energy_cleaned`) covers 1990–2010 across ~140 countries with 52 variables from the EIA/World Bank. Five major economies were selected to represent a range of income levels and energy trajectories. The primary metric — renewable energy as a share of total final energy consumption (TFEC) — includes both modern renewables (wind, solar, hydro) and traditional biomass. A `modern_renewable_pct` column was derived by subtracting the traditional biomass share to allow for richer follow-up analysis, though the main visualization uses the total renewable share to match the source definition. Missing values were present in many columns but not in the key trend variable for the selected countries.

## Reproduce

Run this week's analysis with the following code in your R environment. Make sure you have the necessary system requirements and that you have cloned the repository to your local machine.

```r
setwd("W21")        # set working directory to this week's folder
renv::restore()     # restore locked package environment
quarto::quarto_render("analysis.qmd")
```

**System requirements**

* R version 4.6.x
* quarto cli version `{r} quarto::quarto_version()` or greater
* Optional: Positron version `{r} Sys.getenv("POSITRON_VERSION")` or greater
* Rtools45 for Windows (or equivalent build tools for other OS)

*View [renv.lock](renv.lock) for detailed package information*

### Session Info
```{r}
#| label: dependencies

cat("Quarto version", as.character(quarto::quarto_version()))

sessionInfo()

```
