---
title: "Week 21: Energy Consumption"
description: "Exploring global energy consumption trends using EIA data."
author: "Daniel B. Mason (Cooper)"
date: 2026-05-26
date-modified: last-modified
categories: [energy, environment]
freeze: true
format:
  html:
    toc: true
    toc-location: left
    code-fold: false
    code-tools: true
---

**Data Source**: [TidyTuesday 2026 Data](https://github.com/rfordatascience/tidytuesday/tree/main/data/2026)

```{r}
#| label: setup
#| message: false
source("renv/activate.R")

library(dplyr)
library(ggplot2)
library(tidyr)
library(tidytuesdayR)

source("../utils/themes.R")
source("../utils/palette.R")
source("../utils/helpers.R")

## load data using tidytuesdayR helper function
tt <- tidytuesdayR::tt_load(2026, week = 21)

```

```{r}
#| label: explore data

df <- tt$energy_cleaned

# Dataset dimensions and year range
cat("Rows:", nrow(df), "| Cols:", ncol(df), "\n")
cat("Year range:", min(df$yr), "–", max(df$yr), "\n")
cat("Countries:", n_distinct(df$country_name), "\n")

skimr::skim(df)

```

```{r}
#| label: wrangle data
energy_labels <- c(
  biogas_consumption_tfec_pct = "Biogas",
  geothermal_energy_consumption_tfec_pct = "Geothermal",
  hydro_energy_consumption_tfec_pct = "Hydro",
  liquid_biofuels_energy_consumption_tfec_pct = "Liquid biofuels",
  marine_energy_consumption_tfec_pct = "Marine",
  modern_biomass_energy_consumption_tfec_pct = "Modern biomass",
  solar_energy_consumption_tfec_pct = "Solar",
  traditional_biomass_consumption_tfec_pct = "Traditional biomass",
  waste_energy_consumption_tfec_pct = "Waste",
  wind_energy_consumption_tfec_pct = "Wind"
)

biomass_types <- c("Traditional biomass", "Modern biomass")

wb_url <- paste0(
  "https://api.worldbank.org/v2/country/all/indicator/SP.POP.TOTL",
  "?format=json&per_page=20000&date=2010"
)

pop_2010 <- jsonlite::fromJSON(readr::read_file(wb_url), flatten = TRUE)[[2]] |>
  transmute(
    country_code = countryiso3code,
    population_2010 = as.numeric(value)
  ) |>
  filter(!is.na(country_code), nchar(country_code) == 3, !is.na(population_2010)) |>
  distinct(country_code, .keep_all = TRUE)

country_pop_q <- df |>
  distinct(country_code) |>
  inner_join(pop_2010, by = "country_code") |>
  mutate(
    pop_quartile = ntile(population_2010, 4),
    pop_quartile = factor(
      pop_quartile,
      levels = 1:4,
      labels = c("Q1: smallest populations", "Q2", "Q3", "Q4: largest populations")
    )
  )

non_biomass_by_pop <- df |>
  inner_join(country_pop_q, by = "country_code") |>
  select(country_code, yr, pop_quartile, ends_with("tfec_pct") & !starts_with("renewable")) |>
  pivot_longer(
    cols = ends_with("tfec_pct"),
    names_to = "energy_type",
    values_to = "pct"
  ) |>
  mutate(energy_type = recode(energy_type, !!!energy_labels)) |>
  filter(!(energy_type %in% biomass_types)) |>
  summarize(pct = mean(pct, na.rm = TRUE), .by = c(pop_quartile, yr, energy_type)) |>
  filter(is.finite(pct))

non_biomass_order <- non_biomass_by_pop |>
  summarize(avg_pct = mean(pct, na.rm = TRUE), .by = energy_type) |>
  arrange(desc(avg_pct)) |>
  pull(energy_type)

non_biomass_by_pop <- non_biomass_by_pop |>
  mutate(energy_type = factor(energy_type, levels = non_biomass_order))
```

```{r}
#| label: visualize 

main_plot <- ggplot(non_biomass_by_pop, aes(x = yr, y = pct, fill = energy_type)) +
  geom_area(alpha = 0.8, linewidth = 0.25, color = NA) +
  facet_wrap(~ pop_quartile, ncol = 2, scales = "free_y") +
  scale_fill_brewer(type = "qual", palette = "Paired", name = "Non-biomass source") +
  labs(
    title = "Non-biomass Renewable Share of TFE\n by 2010 Population Quartile",
    subtitle = "Country-level means within each quartile (equal-country weighting)",
    x = NULL,
    y = "Share of TFEC (%)",
    caption = "Population source: World Bank indicator SP.POP.TOTL (2010)."
  ) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_tt()

```

## Visualization

```{r}
#| label: main-plot
#| fig-width: 8
#| fig-height: 6

ggplot2::ggsave(
  "main_plot.png",
  plot   = main_plot,
  width  = 8,
  height = 6,
  dpi    = 300
)

main_plot
```

## Key Findings

- After removing traditional and modern biomass from the stacked areas, hydro remains the dominant non-biomass source across quartiles, while wind and solar increase gradually from low starting levels.

- The smallest-population quartile shows substantially higher and more volatile non-biomass shares than other quartiles, with a notable jump near the end of the series.

- Larger-population quartiles (Q3–Q4) remain lower and flatter overall, suggesting slower aggregate shifts in non-biomass renewable composition under equal-country weighting.

## Methodology

The dataset (`energy_cleaned`) covers 1990–2010 across ~140 countries with 52 variables from the EIA/World Bank. Five major economies were selected to represent a range of income levels and energy trajectories. The primary metric — renewable energy as a share of total final energy consumption (TFEC) — includes both modern renewables (wind, solar, hydro) and traditional biomass. To focus on the former, we excluded traditional biomass and analyzed trends by population quartiles (based on 2010 population) to explore how energy transitions may differ by country size. We used area charts to visualize the composition of non-biomass renewables over time within each population quartile.

## Reproduce

Run this week's analysis with the following code in your R environment. Make sure you have the necessary system requirements and that you have cloned the repository to your local machine.

```r
setwd("W21")        # set working directory to this week's folder
renv::restore()     # restore locked package environment
quarto::quarto_render("analysis.qmd")
```

**System requirements**

* R version 4.6.x
* quarto cli version `{r} quarto::quarto_version()` or greater
* Optional: Positron version `{r} Sys.getenv("POSITRON_VERSION")` or greater
* Rtools45 for Windows (or equivalent build tools for other OS)

*View [renv.lock](renv.lock) for detailed package information*

### Session Info
```{r}
#| label: dependencies

cat("Quarto version", as.character(quarto::quarto_version()))

sessionInfo()

```
